---
title: "Week 2"
output: html_notebook
---

#Preamble

This notebook contains week 2 notes for BIOL432. The lecture includes dplyr and regular expressions. 

#Data science -> making a science out of working with data

Often using big data

Characteristics of big data (the 4 V's)

  * High volume
    - req's standardized approach
  * Variety
    - diff types of data
  * Velocity
    - rapid accumulation of data
  * Compromise verasity
    - variability in quantity of data -> need a way to distinguish & select based on quality
    
=> Data science is the circle of life:

  Hypoth -> Data collection -> Quality control -> Explore data -> Model -> Report -> etc.
  
dplyr -> good for working with 2D data

```{r}
Fallo<- read.csv("FallopiaData.csv")
str(Fallo)
```
^ 2D dataset (i.e. rows & col), each cell contains single value, each var has own col, each row is observation

```{r subset data}
# install.packages("dplyr")
library(dplyr)
# filter()
Pot10<- filter(Fallo,PotNum==10) # searches dataframe Fallo for all cases where PotNum is 10
PotL10<-filter(Fallo,PotNum<10)
PotL10

names(Fallo)[13]<- "Fallopia"
head(Fallo)

rename(Fallo, Total_Biomass = Total)
names(Fallo)

X<- arrange(Fallo, Taxon, Total) # arranges in ascending order
head(X)

X<- arrange(Fallo, Taxon, desc(Total)) # arranges in descending order
head(X)

X<- select(Fallo, PotNum, Scenario, Nutrients, Taxon, Total)
head(X)

X<- select(Fallo, PotNum:Taxon, Total)
head(X) # quick way to shrink data into variables you're interested in

X<- select(Fallo, -Total)
head(X) # removes Total from the dataset

# modify this to exclude PotNum:Taxon, Total
X<- select(Fallo, -c(PotNum:Taxon, Total))
head(X)

Fallo$Total_g<- Fallo$Total/1000
head(Fallo)
select(Fallo, Total_g, Taxon, Scenario, Nutrients, PotNum, everything()) # rearranges the data as written, with everything() meaning leaves what's left as is

X<- mutate(Fallo, Urtica_g = Urtica/1000) # makes new col but using dplyr package
head(X)

X<- transmute(Fallo, Urtica_g = Urtica/1000) # this transforms Urtica & gets rid of everything else

# for summaries...

TrtGrp<- group_by(Fallo,Taxon,Scenario,Nutrients) # makes a tibble... diff than dataframe -> object that dplyr uses to group dataframe
summarize(TrtGrp, Mean=mean(Total), SD=sd(Total), Count=length(Total))

```
If you want to do this a bunch of times, make a pipe

```{r pipe}
X<- group_by(Fallo, Taxon)
X<- group_by(X, Scenario)
X<- group_by(X, Nutrients)
X<- summarize(X, Mean=mean(Total), SD=sd(Total))
arrange(X, desc(Mean))
# sorted mean high to low, but only have 2 means... vs group_by() above has several
# if we make a mistake in this, it's carried on b/c using same object name
# also true if for loop makes a mistake, it's carried on throughout the loop
# doing 3 group_by statements in a row essentially just groups by the last one

# better to do this:
X<- group_by(Fallo, Taxon)
X1<- group_by(X, Scenario)
X2<- group_by(X1, Nutrients)
X3<- summarize(X2, Mean=mean(Total), SD=sd(Total))
arrange(X3, desc(Mean))

Y<- group_by(Fallo, Nutrients)
summarize(Y, Mean=mean(Total), SD=sd(Total))

#pipe

# unweighted mean of nutrients
FalSum<- Fallo%>%
  group_by(Taxon)%>%
  group_by(Scenario)%>%
  group_by(Nutrients)%>%
  summarize(Mean=mean(Total), SD=sd(Total))%>% # don't need to declare data this time b/c taking output as the next input
  arrange(desc(Mean))
# better b/c not making intermittent objects, less error prone

# weighted mean of nutrients
Fallo %>%
  group_by(Taxon, Scenario, Nutrients) %>%
  summarize(Mean=mean(Total)) %>%
  group_by(Scenario,Nutrients) %>%
  summarize(Mean=mean(Mean)) %>% # this will be a mean of means... which isn't always the same as a regular mean, it's a weighted mean (good when diff sample sizes or unbalanced samples)
  group_by(Nutrients) %>%
  summarize(Mean=mean(Mean)) %>%
  arrange(desc(Mean))

Fallo %>%
  group_by(Nutrients) %>%
  summarize(Mean=mean(Total)) %>%
  arrange(desc(Mean))
# this is not identical to above (b/c not mean of means)

```
NA = missing values

a special object in R that is not the same as 0
```{r missing data}
Vec<- c(1,2,3,NA,5,6,7)
Vec

mean(Vec) # this returns NA b/c R doesn't know what to do w/ NA (remove? treat as 0?)

mean(Vec, na.rm=TRUE) # this removes NA values

is.na(Vec) # this returns vector of same length as Vec where TRUE means a missing value

sum(is.na(Vec)) # this returns the number of TRUE's (i.e. the num of missing values)

X<- round(runif(10, min=1, max=nrow(Fallo)), 0) # need to round b/c looking for row numbers
Fallo$Total[X] # these are the values from the randomly selected rows indicated by X
Fallo$Total[X] <- NA # make these rows NA

is.na(Fallo$Total) # instead of Fallo$total==NA 

# to run group_by funtions above with new NA's, need to add na.rm=T to first mean()

```

Naughty data -> not simple, 2D data. can be messy & R can't read. 

Need to use regular expressions

grep() and grepl() -- 'find'

sub() and gsub() -- 'find & replace'

```{r Regular Expressions}
Species<- c("petiolata", "verticillatus", "salicaria", "minor")
Species

grep("a", Species) # outputs an index vector (i.e. there is an a in the first, second, & third cell)
grepl("a", Species) # outputs boolean vector

sub("a", "X", Species) # only replaces the first a
gsub("a", "X", Species) # replaces all a's b/c evalutates each cell independently 

# these are both case sensitive 

# \ is an escape character -> what's coming after the slash is diff/ you need to treat it differently
# \w -- all letters & digits ('word')

gsub("w", "X", "...which 1-100 words get replaced?") # every w is replaced
gsub("\\w", "X", "...which 1-100 words get replaced?") # replaces every letter & number (not punctuation)

# \s is blank spaces
gsub("\\s", "X", "...which 1-100 words get replaced?") # replaces every space


# \t is a tab character
gsub("\\t", "X", "...which 1-100 words get replaced?") # good for tabbed data

# \d for digits
gsub("\\d", "X", "...which 1-100 words get replaced?") # replaces every number

# \D for non-digits
gsub("\\D", "X", "...which 1-100 words get replaced?") # replaces everything that is not a number

# \r and \n are new line chacters
# good for making output string w/ seprate lines or a file w/ multiple lines that you want to use regular expressions on
X<- gsub("\\s", "\n", "...which 1-100 words get replaced?") # note: always just one \ for n
cat(X)
write.csv(X, "test.csv")

# | -- or 
gsub("w|e", "X", "...which 1-100 words get replaced?") # replaces either a w or an e

gsub(".", "X", "...which 1-100 words get replaced?") # replaces everything... "." is a special character that replaces everything except the new line character

# *, ?, + to modify searches
sub("\\w+", "X", "...which 1-100 words get replaced?") # + means one or more matches, replaces the whole word
gsub("\\w+", "X", "...which 1-100 words get replaced?") # replaces all words

# * means zero or more matches
sub(".*", "X", "...which 1-100 words get replaced?") # replaces any number of any character

# {} match particular numbers of matches
gsub("\\w{5}", "X", "...which 1-100 words get replaced?") # replaces words with 5 letters or the first 5 letters of the word
gsub("\\w{2}", "X", "...which 1-100 words get replaced?") # replaces every 2 letters in a word
gsub("\\w{2,5}", "X", "...which 1-100 words get replaced?") # replaces both 2 & 5 letters?

# List options w/ []
gsub("[abc]", "X", "...which 1-100 words get replaced?") # replaces a or b or c
gsub("[a-h]", "X", "...which 1-100 words get replaced?") # replaces a range of numbers

# ^ and $ for start/end of line
sub("^s", "X", Species) # only s as the first letter is replaced

# Capture with ()
gsub("(\\w)\\w+", "\\1", "...which 1-100 words get replaced?") # replaces the second letter (keeps the first), replaces what's in the first set of brackets with just the first letter

gsub("(\\w)(\\w+)", "\\2\\1", "...which 1-100 words get replaced?") # takes the first letter & puts it on the end
```

# Review

## Data Science

Data science is a relatively new field of study that merges comp sci & stats w/ other domains of knowledge (i.e. business, bio, psych). Data science as a discipline has grown in popularity in response to the rapid rate of increase in data collection and publication. 


Data science often invoves 'Big Data', which doesn't have a strict quantitative def'n but will usually have one or more of the following characteristics:

(The 4 V's of Big Data):
  * High Volume 
    - lrg file sizes w/ lots of observations
  * Wide Variety
    - lots of diff types
  * High Velocity
    - accumulating at a high rate
  * Compromised Veracity
    - variable quality that must be dealt otherwise downstream analyses will be compromised
    
## 2D Data Wrangling


```{r Review Part 1}
library(dplyr)
Fallo<- read.csv("FallopiaData.csv")
str(Fallo)
```
This file is an example of a 2-D data set, which is common in bio. 2D datasets have the familar row x col layout used by spreadsheet programs like Excel or Google Sheets. There are some exceptions, but data in this format should typically follow 3 rules:

  1. Each cell contains a single value
  2. Each variable must have its own col
  3. Each observation must have its own row
  
Making sure your data is arranged this way will usually make it much easier to work with. 

```{r Review Part 1 contd}
# subset observations based on value
Pot1<- filter(Fallo, PotNum==1)

# change col names with [] index
X<- Fallo
names(X)
names(X)[12]<- "Total_Biomass"
names(X)

# change col names with rename()
X<- rename(Fallo, Total_Biomass = Total)
names(X)

# sort rows based on col's with arrange()
X<- arrange(Fallo, Taxon, Total)
head(X)

# use desc() with arrange() to change to descending order
X<- arrange(Fallo, Taxon, desc(Total))
head(X)

# select a subset of col's with select()
X<- select(Fallo, PotNum, Scenario, Nutrients, Taxon, Total)
head(X)

# can also use ":" to select range of col's
X<- select(Fallo, PotNum:Taxon, Total)
head(X)

# exclude col's with -
X<- select(Fallo, -(PotNum:Taxon), -Total)
head(X)

# use everything() with select() to rearrange w/o losing any
X<- select(Fallo, Taxon, Scenario, Nutrients, PotNum, Pct_Fallopia, everything())
head(X)

# to make a new col in base R, use $
X<- Fallo
X$UrtSil<- X$Urtica+X$Geranium

# use mutate() in dplyr to make new col
X<- mutate(Fallo, UrtSil = Urtica + Geranium)
head(X)

# use transmute() to retain new col's and delete everything else
X<- transmute(Fallo, UrtSil = Urtica + Geranium)
head(X)

# use summarize() + group_by() to quickly summarize data 
TrtGrp<- group_by(Fallo, Taxon, Scenario, Nutrients)
summarize(TrtGrp, Mean=mean(Total), SD=sd(Total))

# %>% (pipe) is useful for combining operations w/o creating new objects
## instead of:
X<- group_by(Fallo, Taxon)
X<- group_by(X, Scenario)
X<- group_by(X, Nutrients)
X<- summarize(X, Mean=mean(Total), SD=sd(Total))
arrange(X, desc(Mean))

## use:
Fallo %>%
  group_by(Taxon) %>%
  group_by(Scenario) %>%
  group_by(Nutrients) %>%
  summarize(Mean=mean(Total), SD=sd(Total)) %>%
  arrange(desc(Mean))
```

## Missing Data

The R language uses a special object, NA, to denote missing data

When a function is run on a vector or other object containing NA, the function will often return NA or give an error message. 

This is by design, because it is not always clear what NA means. Manny functions in R include an na.rm parameter that is set to FALSE by default. Setting it to true tells the function to ignore the NA

```{r Review Part 2}
Vec1<- c(1,2,3,NA,5,6,7)
Vec1

mean(Vec1, na.rm=T)

# NA is different than 0
Vec2<- c(1,2,3,0,5,6,7)
mean(Vec2, na.rm=T)

# use is.na() to check for missing values
## generate test dataset by randomly selecting 10 rows & replacing 'Total' w/ NA
X<- round(runif(10, min=1, max=nrow(Fallo)),0)
Fallo$Total[X]<- NA
Fallo$Total
## check for missing values with is.na()
is.na(Fallo$Total)
### the output is a boolean variable which has handy properties

# boolean variable can be used as an index 
## see which pots have missing 'Total' values
Missing<- is.na(Fallo$Total)
Fallo$PotNum[Missing]

## can also count missing values
sum(is.na(Fallo$Total))
```

## Naughty Data

Naughty data contain the same info as a standard row x col (i.e. 2D) dataset but break the rules outlined above:

  1. Each cell cotnains a single value
  2. Each variable must have it's own col
  3. Each observation must have its own row
  
Naughty data is hard to fix but regular expressions make it easier...

## Regular Expressions

Regular expressions 

  * aka 'regex' and 'regexp'
  * a sort of find-and-replace for nerds
  * powerful data tool
  * requires patience and lots of practice 
  
Four Basic Regex Functions in R:

  `grep() and grepl()` are equivalent to 'find' in a word processor
  
  * General form: `grep("find this", in.this.object)`
  
  `sub() and gsub()` are equivalent to 'find and replace'
  
  * General form: `grep("find this", "replace with this", in.this.object)`
  
Three Advanced Regex Functions in R:

  `regexpr()` provides more detailed info about the first match
  
  `gregexpr()` provides more detailed results about all matches
  
```{r Review Part 3}
# create a simple data frame of species names
Species<- c("petiolata", "verticillatus", "salicaria", "minor")
print(Species)

# grep() returns cell addresses matching query
grep("a", Species)

# grepl() returns T/F associated with query
grepl("a", Species)

# sub() replaces first match (in each cell)
sub("l", "L", Species)

# gsub() replaces all matches
gsub("l", "L", Species)
```

Wildcards

  `\` escape character
  
  `\\` double escape, usually needed b/c first `\` used to escape sepcial character in R and `\` is itself a special character that needs to be escaped to pass through the function
  
```{r Review Part 3 contd}
# Escape characters

## \\w = all letters and digits (aka 'words')
gsub("w", "*", "...which 1-100 words get replaced?")
gsub("\\w", "*", "...which 1-100 words get replaced?")

## \\W = non-word and non-number (inverse of \\w)
gsub("\\W", "*", "...which 1-100 words get replaced?")

## \\s = spaces
gsub("\\s", "*", "...which 1-100 words get replaced?")

## \\t = tab character (usful for tab-deliminated data)
gsub("\\t", "*", "...which 1-100 words get replaced?")

## \\d = digits
gsub("\\d", "*", "...which 1-100 words get replaced?")

## \\D = non-digits
gsub("\\D", "*", "...which 1-100 words get replaced?")

## \\r = carriage return
## \\n = newline character

# Special characters

## | = 'or'
gsub("w|e", "*", "...which 1-100 words get replaced?")

## . = any character except new line
gsub(".", "*", "...which 1-100 words get replaced?")

## search for a period with \\.
gsub("\\.", "*", "...which 1-100 words get replaced?")

# use *, ?, +, and {} for more complicated searches

## + = 1 or more matches
sub("\\w+", "*", "...which 1-100 words get replaced?")
gsub("\\w+", "*", "...which 1-100 words get replaced?")

## ? is a 'lazy' search (find the first occurrence)
sub("\\w?", "*", "...which 1-100 words get replaced?")
gsub("\\w?", "*", "...which 1-100 words get replaced?")

## * is a 'greedy' search (find the last occurrence)
sub("\\w*", "*", "...which 1-100 words get replaced?")
gsub("\\w*", "*", "...which 1-100 words get replaced?")

## {n,m} = between n to m matches
gsub("\\w{3,4}", "*", "...which 1-100 words get replaced")

## {n} = exactly n matches
gsub("\\w{3}", "*", "...which 1-100 words get replaced")

## {n,} = n or more matches
gsub("\\w{4,}", "*", "...which 1-100 words get replaced")

# List range of options using []
gsub("[aceihw-z]", "*", "...which 1-100 words get replaced")

# find everything in square brackets occuring 1 or more times
gsub("[aceihw-z]+", "*", "...which 1-100 words get replaced")

# ^ start of line
## find species starting with s
grep("^s", Species)
## IMPORTANT: ^ also 'negates' when used within []
### find species containing any letter other than s
grep("[^s]", Species)
### replace every letter except s
gsub("[^s]", "*", Species)

# $ end of line
## find species ending with a
grep("a$", Species)

# capture text using () and reprint using \\1, \\2, etc
## replace each word with its first letter
gsub("(\\w)\\w+", "\\1", "...which 1-100 words get replaced?")
## pull out only the numbers and reverse their order
gsub(".*([0-9]+)-([0-9]+).*", "\\2-\\1", "...which 1-100 words get replaced?")
## reverse first two letters of each word
gsub("(\\w)(\\w)(\\w+)", "\\2\\1\\3", "...which 1-100 words get replaced?")
```

# Git Version Control Video Questions

  1. Briefly explain the differences between a 'save', 'commit', 'branch', and 'merge' in Git
  
save = save that version over your old version

commit = save recent changes with diffs to git

branch = create a second route of commits without altering the original commit seq

merge = join branches to a single commit 

  2. Why are commits more advantageous than regularly saving your files?
  
Commits track changes (diffs) for version control. Enhances collaboration b/c changes from each collaborator are easy to follow. Allows one to return to an older version to restart a process or create a second version (i.e. branch). 

  3. What is the .gitignore file used for?

This removes files from the commit 

  4. What are ‘push’ and ‘pull’ used for?
  
push = refresh current commits to github 

pull = take commits from someone else's github

  5. What’s the difference between Git and GitHub?
  
Git is the program in R allowing version control. Github is a repository allowing versions to be shared publically.

  6. What is your GitHub username?

haybracky

#Practice Exercises

Consider a vector of email addresses scraped from the internet:

  * robert ‘dot’ colautti ‘at’ queensu ‘dot’ ca
  * chris.eckert[at]queensu.ca
  * lonnie.aarssen at queensu.ca
  
Use regular expressions to convert all email addresses to the standard format: name@queensu.ca

```{r}
Rob1<- "robert ‘dot’ colautti ‘at’ queensu ‘dot’ ca"
Rob2<- gsub("\\W", " ", Rob1)
Rob3<- gsub("dot", ".", Rob2)
Rob4<- gsub("at", "@", Rob3)
Rob5<- gsub("\\s", "", Rob4)
Rob5

c<- gsub("at", "@", chris)
c1<- gsub("\\[|\\]", "", c)
c1

lon<- "lonnie.aarssen at queensu.ca"
lon1<- sub("(\\s)at(\\s)", "@", lon)
lon1
```


# Quiz

```{r}
words<- "Write a line of R code that will switch the first and last letter of every word in this sentence if the word is >3 letters long"
gsub("(\\w+)(\\w)(\\w)(\\w+)", "\\4\\2\\3\\1", words)
```

