---
title: "Assignment 2"
author: "Hayley Brackenridge"
date: '2018-01-16'
output:
  pdf_document: default
  html_document: default
  word_document: default
---

#Preamble

This Notebook is a compilation of basic R functions formatted for future use in scientific paper publication. It includes R details, package loading, a description of variables, data file manipulation, summary tables, and confidence intervals. The Notebook uses the dataset published by Sandau et al. 2016 collected from 2002-2009 on plant ecosystem diversity and biomass production. This Notebook was created by Hayley Brackenridge on January 16th, 2018 for BIOL 343 Assignment #2 at Queen's University. 

##R Details

  * File creation date`r Sys.Date()`
  * `r R.version.string`
  * `ggplot2` package `r packageVersion("ggplot2")`
  * `dplyr` package `r packageVersion("dplyr")`
  * `Publish` package `r packageVersion("Publish")`

##Load Packages
```{r Packages}
library(ggplot2)
library(dplyr)
library(Publish)
```

##The Data

Data uploaded from DRYAD

Variables:
  * PlotID = unique plot identification
  * Year = year data was collected (either 2008 or 2009)
  * Field = field number data was collected from
  * Treat = fencing treatment; categorical variable with categories: C = control; PE = predator exclusion; PHE = predator and herbivore exclusion
  * SDiv = number of sown plant species 
  * LeftS = number of sown species still present
  * ColS = number of colonising species (predators & herbivores?)
  * Tot_S = total species richness 
  * W_m2 = total above ground biomass in g*m-2

```{r my data}
dat<-read.csv("BB_all_4_SimilMatrices_Dryad.csv")
dat1<-dat[c(1:3, 5:10)]
write.csv(dat1, file="dataset1.csv")
dat2<- dat1%>% mutate(log10(W_m2), Ts= Tot_S, Bs= W_m2, 10*ColS)
dat2
write.csv(dat2, file="dataset2.csv")
dat3<-subset(dat2, Treat=='8mm')
dat3
write.csv(dat3, file="dataset3.csv")
```

```{r summary}
sum<-dat3%>%
  summarise(n=n(),mean(Bs),median(Bs),min(Bs),max(Bs),sd(Bs))
sum
CI<-ci.mean(Bs~Year, data=dat3)
CI
plot(CI)
```
##Interpretation

The above graph shows the 95% confidence interval for total above ground biomass in the 8mm fence treatment only over both 2008 and 2009. This suggests that there is a 95% chance the true mean of total above ground biomass for the 8mm fence treatment lies in this range. 
