---
title: "Assignment 3"
author: "Hayley Brackenridge"
date: '2018-01-24'
output:
  html_document:
    df_print: paged
---

#Preamble

This notebook contains assignment 3 for BIOL 343. A loop will be created to simulate a random sample of 2 species of turtles. The loop will be run for samples of 50 and 5. A ND of the sample means will be plotted and compared. 

```{r generate data}
# generate data from ND
## (n, mean, SD)
### species A
dat1<-rnorm(900, 82, 2)
### species B
dat2<-rnorm(900, 80, 2)
# set up variable to be empty
dat3<- 0
dat4<- 0
```

```{r loop 1}
# make a loop, i, that will start at 1 & continue up to 100
for (i in 1:100){
  X<- sample(dat1, 50)
  Y<- sample(dat2, 50)
  dat3[i]<- mean(X)-mean(Y)
  }
dat3
# dat3 contains the difference of the means generated from a random sample of 50 individuals from species A and B
```

```{r probability 1}
# find the probability that species A is bigger than species B
## calculates how many differences are greater than or equal to zero
data<- as.data.frame(dat3)
filter(data, dat3 >= 0)
# 100% values greater than or equal to 0
```

```{r loop 2}
# make a loop, i, that will start at 1 & continue up to 100
for (i in 1:100){
  X<- sample(dat1, 5)
  Y<- sample(dat2, 5)
  dat4[i]<- mean(X)-mean(Y)
  }
dat4
# dat4 contains the difference of the means generated from a random sample of 5 individuals from species A and B
```

```{r probability 2}
# find the probability that species A is bigger than species B
## calculates how many differences are greater than or equal to zero
data2<- as.data.frame(dat4)
filter(data2, dat4 >= 0)
# 97% values greater than or equal to 0
```

```{r distributions}
# histogarm of differences from loop 1
hist(dat3)
# qqplot of differences from loop 1
qqnorm(dat3)
# the distributions are roughly normal
```


##Explanation

Populations of two species, A and B are generated from a normal distribution contianing 900 individuals, one with a mean weight of 82g and the other of 80g with standard deviations of 2. These are named dat1 and dat2. dat3 is set up to be an empty variable to use later. 

Loop 1 is generated whereby a sample of 50 randomly selected individuals from each population is selected, their mean calculated, and the difference between the means calculated. The loop repeats this sampling 100 times and stores the differences as a dataset in dat3. 

The probability of species A being bigger than species B is calculated. For loop 1, the probability is 100%. 

Loop 2 is generated whereby a sample of 5 randomly selected individuals from each population is selected, their mean calculated, and the difference between the means calculated. The loop repeats this sampling 100 times and stores the differences as a dataset in dat4. 

The probability of species A being bigger than species B is calculated. For loop 2, the probability is 97%. 

The probabilities generated by the two loops are different because of the different sample sizes. As sample size increases, the amount of variation decreases as the sample becomes more representative of the true population. 