---
title: "Week 1 Class Notes"
output: html_notebook
---

Date: 08/01/19

#Preamble 

This notebook contains class notes for BIOL432 Week 1

#Content

Project files help contain files and make them easy to share

```{r check wd}
getwd()
```
When you open a proj file, it will default to your directory, regardless of where it's saved

```{r math commands}
10+2
10*2
10-2
10/2
10^2
abs(-10) # absolute value
sqrt(10-1)
log(10)
log10(10)
?log # get help with the function log
log(64,4) # log base 4 of 64
exp(4)
sin(pi/2)
acos(-1)
tan(0)
atan(0) # inverse tangent
round(pi,3) # rounds pi to 3 digits
floor(pi) # rounds pi down
ceiling(pi) # rounds pi up
signif(pi,digits=2) # sets num of sig dig for pi
```

Important that the code is readable & interpretable 

```{r arranging data}
log(64,
    4) # functions can span more than one line
# helps organize data in a readable way
```

## Logic Operators

```{r logic operators}
1>2
1<2
1<=2
1=1 # this does not work b/c = used to define objects
test<- 0 # defines the object test as 0 ... in other words puts 0 in the object test
test=9 # also puts 9 in the object test. Both methods work
1==1 # this is used to evaluate 
1!=1 # not equal to
1==2 | 1==1 # asking if one is equal to 2 OR 1
1==2 & 1==1 # asking if one is equal to 2 AND 1
```

```{r make group of numbers}
numbers<- c(1,2,3,4,5,6,7,8) # makes a vector called numbers
numbers
4 %in% numbers # is 4 in the vector numbers?
```

```{r selecting numbers}
runif(10,min=0,max=1) # selects 10 random numbers betn 0 and 1 from a uniform distribution
rnorm(10,mean=0,sd=1) # selects 10 random numbers from a normal distribution with mean 0 and standard deviation of 1
rpois(10,lambda = 10) # selects 10 random numbers from a poison distribution with lambda = 1
rbinom(10,size=1,prob=0.5) # this is like flipping a coin with 2 sides once
rbinom(10,size=10,prob=0.5) # this is like doing it 10 times... 5/10 heads, 4/10 heads, 2/10 heads, etc. 
```

can be hard to tell if output numbers b/c of randomness or b/c of error
trial this by using pseudosampling

```{r pseudosampling}
set.seed(3)
runif(5)
# comps can't generate truly random nums. by setting a seed, you force the comp to use the same seed. you will always get the same set of random nums by using that same seed
```

```{r concatenate}
# there are 3 types of data: integer, float, string
  # float = num w/ decimal, real number
  # string = everything else

#combine these
c(1,2,3.14) # automatically converts integers to floats

c(1,2,3.14,"test") # automatically converts all to strings
c(1,2,3.14,lkjava) # treats it as an object rather than a string, error b/c object not defined

c(1:100) # outputs every integer btwn 1 & 100
c(-10:10) # same but using -ve nums

X<- "string"
X
print(X) # needed for loops
paste (X) # good for combining things
Xvec<- c(X,1:10,"E","Computation",100:50)
Xvec
# note: all objects created are listed under Global Enviornment
  # this is helpful for troubleshooting :)

# num corresponding to element in vector in [ ]
Xvec[55] # retrieves 55th element in the vector Xvec
```

```{r matrix}
Xmat<- matrix(Xvec, nrow=8) # number of rows (nrow) needs to be multiple of data length (Xvec)
Xmat # this is an 8 x 8 matrix
# [x, ] = row num   [ ,x] = col num
Xmat[5,] # pulls out everything in the 5th row
Xmaat[5,3] # pulls out element in 5th row, 3rd col
```

```{r List}
MyList<- list(name="biol432",potpourri=Xvec,numbers=1:10)
MyList # this object has diff components (name, potpourri, numbers)

MyList$numbers # subsets list

# general form of lm:
#lm(Y~X)
testmod<- lm(rnorm(10)~rnorm(10))
names(testmod) # this is just a list

testmod$coefficients
```

This is useful for reproducibility & replication 

```{r Slice & Extract}
MyList[3] # slice

MyList[[3]] # extract -> this does not keep the numbers w/ the name of the list

MyList[[3]]*10
MyList[3]*10 # this only works with extract b/c slice takes the title of the list so it doesn't treat it as numbers
testmod[[1]]
```

```{r Paste}
paste("Hello world!")
paste("Hello","World","!") # automatically puts spaces btwn each element
paste(1:10) # keeps these as a string

paste("Hello",1:10) # Hello is vector length 1, 1:10 is vector length 10 ... takes element from each 10 numbers & the next element of Hello but only length 1 so all say Hello

paste(1:2,1:10) # takes vector of 2 num & vector of 10 num & pastes together
```

```{r working w/ data}
Fallopia<- read.csv("FallopiaData.csv")
Fallopia
head(Fallopia)
names(Fallopia)
tail(Fallopia)
str(Fallopia)
dim(Fallopia)
nrow(Fallopia)
ncol(Fallopia)

# combine data of Symphytum & Silene biomass
Fallopia$combine<- Fallopia$Symphytum+Fallopia$Silene
# could also do...
Fallopia$combine<- Fallopia["Symphytum"]+Fallopia["Silene"]
# don't use col or row num b/c if data set changes, these change & you wouldn't notice

levels(Fallopia$Nutrients)

LowNut<- subset(Fallopia, Nutrients=="low") # subset data
write.csv(LowNut,"Subset.csv") # this makes the subset as a new csv file saved in the wd

names(Fallopia)

# want to know total biomass in high vs low
aggregate(Total~Nutrients, data=Fallopia, mean) # takes col Total & groups by Nutrients in data set Fallopia & calculates mean
aggregate(Total~Nutrients*Scenario, data=Fallopia, mean) # also groups by scenario
aggregate(Total~Nutrients, data=Fallopia, sd) # gives standard deviation

tapply(Fallopia$Total, list(Fallopia$Nutrients), mean) # gives same results 

```

```{r Flow Control}
X<- 10
X<- 100000

if(X > 100) {
  print("X > 100")
} else {
  print("X <= 100")
}

for(i in 1:X) {
  print(i)
} ## takes object i & makes it 1 & does what's in {}, then comes back & makes i 2, etc.

for(i in Xvec){
  print(i)
}

# write for loop to cycle thru 100 nums
# if num is even, print num
# if num is odd, don't print
for(i in 1:100){
  if (i%%2==0) {
    print(i)
  }
}

set.seed(3) # setting seed so randomly selected numbers from uniform distribution will always be the same
for (i in runif(10,min=-1,max=1)){
if (i>0){
print(i)
}
}
# this loop selects 10 numbers from the uniform distribution from seed 3 & only prints when positive 

# print isn't necessary if manipulating / creating vectors based on other vectors

```

